Informative Oceanographic Sampling using Excursion Probabilities for Multivariate Random Fields

### 2019-09-26 main_new.tex

Our objective is to leverage statistical tools to characterize a river plume, focusing on spatial separation of cold freshwater from the river and warmer saline waters of the fjord. Excursion sets (ES) and excursion probabilities (EP) are useful starting points to measure the ability to classify water masses. The goal of experimental sampling is to improve this characterization, as quantified via changes in the EPs.

@@

Calculations of the EPs and the IBV require a model specification. We use Gaussian modeling, which provides a practical and popular probabilistic approach to modeling environmental data, allowing a closed form
evaluation of the IBVs, as derived in Section \ref{sec:sur}.

@@

The Gaussian model is used to represent the bivariate random fields of temperature and salinity. At location $\bx \in \mathcal{M}$, the prior bivariate distribution is:
\begin{equation}\label{gp_bar}
  \begin{bmatrix}\xi_T(\bx) \\
    \xi_S(\bx) \end{bmatrix}
 \sim N \left( 
\begin{bmatrix} \mu_{T}(\bx)\\
\mu_{S}(\bx)
\end{bmatrix},\begin{bmatrix}
\sigma_{T}^2(\bx) & \sigma_{T}(\bx) \sigma_S(\bx) \gamma(\bx)  \\
\sigma_{T}(\bx) \sigma_S(\bx) \gamma(\bx)  & \sigma_{S}^2(\bx) 
\end{bmatrix}
\right),
\end{equation}
where the notation refers to Gaussian (Normal) distributed variables
specified by the mean vector and covariance matrix. For the application
of tracking river plumes, the model parameters are specified from a short preliminary survey. The mean varies with location, capturing the variability near the river mouth, while the variance parameters and correlation between salinity and temperature are assumed to be constant for all locations and within each water mass.

For the spatial dependence among variables, we assume a separable correlation function with the same decay for salinity and temperature,
which is not unrealistic for our case with a river plume as the decay is related to the common water masses (Sect \ref{sec:case_study}). We have:
\begin{equation}\label{gp_corr}
\mbox{Cov}(\xi_i(\bx),\xi_j(\bx')) = \rho(\bx,\bx') \sigma_i \sigma_j (\gamma +(1-\gamma )\delta_{ij}), \\
\end{equation}
for $i,j \in {T,S}$ and $\delta_{ij}=1$ if $i=j$ and $\delta_{ij}=0$
otherwise. Moreover, we assume stationary isotropic random fields a priori where the
correlation function $\rho(\bx,\bx')$ solely depends on $\bx$ and
$\bx'$ via a Euclidean distance measure $h=\sqrt{||\bx-\bx'||^2}$. With
data and prior knowledge one could possibly fit and estimate
parameters of non-stationary or non-isotropic covariance functions and
more complex multivariate spatial covariance functions
\citep{gneiting2010matern,genton2015cross}; however that is left for
future work.

We discretize the spatial domain
$\mathcal{M}$ to a set of $n$ grid locations
$\mathcal{M}_g = \{\bx_i, i=1,\ldots,n \}$, where each cell has area $\Delta$. The
grid is used for the waypoint graph setting the possible design locations. It is also used when solving integrals expressions in Eq. (\ref{measVAR}) and
(\ref{sur}) which are then approximated by sums over all grid cells.  Using vector
notation, we denote the Gaussian temperature and salinity variables on
the grid and its density function by: 
\begin{equation}\label{prior}
    \bxi = (\xi_T(\bx_1),\xi_S(\bx_1),\ldots,\xi_T(\bx_n),\xi_S(\bx_n))^t, \hspace{3mm}
    \bxi  \sim  N(\bmu, \bSigma) %\pi (\bxi) & =& \frac{1}{(2\pi)^{n}|\bSigma|^{\frac{1}{2}}}e^{-\frac{1}{2}(\bxi -\bmu)^T\bSigma^{-1}(\bxi-\bmu)}.
\end{equation}
%Here, $\pi(\bxi)$ is the joint density function of length $2 n$ vector $\bxi$, and
The length $2 n$ vector $\bmu$ holds the expected values of
temperature and salinity at the grid locations, while the
$2n \times 2n$ matrix $\bSigma$ holds the
variance-covariances of temperature and salinity variables at all grid
locations.

@@

Data is gathered by an AUV that can measure temperature and salinity
at a subset of locations on the defined grid. This can be done in a
static manner, where the survey design is pre-planned, or in a
sequential way with several rounds of data gathering. For the latter,
a round of data is gathered first, and then assimilated to get an
updated model. Then the updated model is used to design and select the
next round of measurements, and so on. We describe the model for one
round of data and the Gaussian formula for a single update. In Section
\ref{sec:heuristics} we discuss sequential data gathering in the
context of adaptive design of AUV measurements.

We denote the data at $n_y$ locations by
$\by=(y_{T,1},y_{S,1},\ldots,y_{T,n_y},y_{S,n_y})^t$, representing the
temperature and salinity measurements gathered at the first round of
nodes in the grid. The conditional model for the data, given the true temperature
and salinity, is described by: 
\begin{equation}\label{likelihood}
\by | \bxi \sim N( \bG \bxi, \bR), %\pi(\by | \bxi)= \frac{1}{(2\pi)^{m}|\bR|^{\frac{1}{2}}}e^{-\frac{1}{2}(\by -\bG\bxi)^T\bR^{-1}(\by-\bG\bxi)},
\end{equation}
where $\bG$ is an $2n_y \times 2n$ matrix holding '$1$'s on the
measurement indices on the grid and '$0$' otherwise. The size
$2n_y \times 2n_y$ covariance matrix $\bR$ contains diagonal entries
$r^2_T$ and $r^2_S$ which are measurement error variances of
temperature and salinity observations.

The GP mean and covariance in Eq. (\ref{prior}) are updated when more information is available. The
updated distribution for temperature and salinity variables on the
spatial grid, given survey data $\by$, is Gaussian with mean and
covariance: 
\begin{eqnarray}\label{gp_upd}
  \bm &=& \bm(\by) = \bmu+\bSigma \bG^t (\bG \Sigma \bG^t+\bR)^{-1}(\by-\bG \bmu),  \\
  \bA &=& \bSigma - \bSigma \bG^t (\bG \bSigma \bG^t+\bR)^{-1} \bG
          \bSigma.\nonumber
\end{eqnarray}

@@

% The conditional mean $\bm$ will be important in the following
% derivation in Section \ref{sec:sur}.
The calculation of the design criterion in Eq. (\ref{sur}) requires
the marginal distribution of the data which is defined over $\bxi$ to
be $\by \sim N( \bG \bmu,\bG \bSigma \bG^t+\bR)$.  However, it turns
out that a simplified form of the calculation is possible because the
conditional mean in Eq (\ref{gp_hat}) is a linear (affine) function of
the data $\by$ and the covariance is not a function of the data.
Before knowing the data, the distribution of the conditional mean is
\begin{equation}\label{distxi} \bm \sim N(\bmu , \bPsi), \hspace{3mm}
  \bPsi=\bSigma \bG^t (\bG \bSigma \bG^t+\bR)^{-1} \bG \bSigma.
\end{equation} 
When considering only grid location $\bx$, we can
extract elements from its mean vector and covariance matrix to get the
bivariate distribution \begin{equation}\label{dist_mxi} 
  \bm(\bx) \sim N \left( \begin{bmatrix}
      \mu_{T}(\bx) \\
      \mu_{S}(\bx) \end{bmatrix}, \begin{bmatrix}
      \psi^2_{T}(\bx) & \psi_{T,S}(\bx)\\
      \psi_{T,S}(\bx) & \psi^2_{S}(\bx) \end{bmatrix} \right),
\end{equation}
which is important for the derivation of the closed form solution of the design criteria.

>>>>

Our objective is to leverage statistical tools to characterize a river plume, focusing on spatial separation of cold freshwater from the river and warmer saline waters of the fjord. ES and EPs are useful starting points to measure the ability to classify water masses. The goal of experimental sampling is to improve this characterization, as quantified via changes in the EPs.

@@

Calculations of the EPs and the IBV require a model specification. We use Gaussian modeling, which provides a practical and popular probabilistic approach to modeling environmental data, and allows a closed form
evaluation of the IBVs as derived in Section \ref{sec:sur}.

@@

The Gaussian model is used to represent the bivariate random fields of temperature and salinity. At location $\bx \in \mathcal{M}$, the prior bivariate distribution is
\begin{equation}\label{gp_bar}
  \begin{bmatrix}\xi_T(\bx) \\
    \xi_S(\bx) \end{bmatrix}
 \sim N \left( 
\begin{bmatrix} \mu_{T}(\bx)\\
\mu_{S}(\bx)
\end{bmatrix},\begin{bmatrix}
\sigma_{T}^2(\bx) & \sigma_{T}(\bx) \sigma_S(\bx) \gamma(\bx)  \\
\sigma_{T}(\bx) \sigma_S(\bx) \gamma(\bx)  & \sigma_{S}^2(\bx) 
\end{bmatrix}
\right),
\end{equation}
where the notation refers to Gaussian (i.e. normal) distributed variables
specified by the mean vector and covariance matrix. For the application
of tracking river plumes, the model parameters are specified from a short preliminary survey. The mean varies with location, capturing the variability near the river mouth, while the variance parameters and correlation between salinity and temperature are assumed to be constant for all locations and within each water mass.

For the spatial dependence among variables, we assume a separable correlation function with the same decay for salinity and temperature,
which is consistent with expectations for  a river plume where the decay is related to the common water masses (Sect \ref{sec:case_study}). In the resulting function, we have
\begin{equation}\label{gp_corr}
\mbox{Cov}(\xi_i(\bx),\xi_j(\bx')) = \rho(\bx,\bx') \sigma_i \sigma_j (\gamma +(1-\gamma )\delta_{ij}), \\
\end{equation}
where for $i,j \in {T,S}$ and $\delta_{ij}=1$ if $i=j$, and $\delta_{ij}=0$
otherwise. Moreover, we assume stationary isotropic random fields a priori where the
correlation function $\rho(\bx,\bx')$ solely depends on $\bx$ and
$\bx'$ via a Euclidean distance measure $h=\sqrt{||\bx-\bx'||^2}$. With
data and prior knowledge, one could possibly fit and estimate
parameters of non-stationary or non-isotropic covariance functions and
more complex multivariate spatial covariance functions
\citep{gneiting2010matern,genton2015cross}; however, that is left for
future work.

We discretize the spatial domain
$\mathcal{M}$ to a set of $n$ grid locations
$\mathcal{M}_g = \{\bx_i, i=1,\ldots,n \}$, where each cell has area $\Delta$. The
grid is used for the waypoint graph setting the possible design locations. It is also used when solving integrals expressions in Eq. (\ref{measVAR}) and
(\ref{sur}), which are then approximated by sums over all grid cells.  Using vector
notation, we denote the Gaussian temperature and salinity variables on
the grid and its density function by the formula: 
\begin{equation}\label{prior}
    \bxi = (\xi_T(\bx_1),\xi_S(\bx_1),\ldots,\xi_T(\bx_n),\xi_S(\bx_n))^t, \hspace{3mm}
    \bxi  \sim  N(\bmu, \bSigma) %\pi (\bxi) & =& \frac{1}{(2\pi)^{n}|\bSigma|^{\frac{1}{2}}}e^{-\frac{1}{2}(\bxi -\bmu)^T\bSigma^{-1}(\bxi-\bmu)}.
\end{equation}
%Here, $\pi(\bxi)$ is the joint density function of length $2 n$ vector $\bxi$, and
The length $2 n$ vector $\bmu$ contains the expected values of
temperature and salinity at the grid locations, while the
$2n \times 2n$ matrix $\bSigma$ contains the
variance-covariances of temperature and salinity variables at all grid
locations.

@@

Data is gathered by an AUV that can measure temperature and salinity
at a subset of locations on the defined grid. This can be done in a
static manner (where the survey design is pre-planned), or in a
sequential way with several rounds of data gathering. For the latter,
a round of data is gathered first, and then assimilated to get an
updated model. Then the updated model is used to design and select the
next round of measurements, and so on. We describe the model for one
round of data and the Gaussian formula for a single update. In Section
\ref{sec:heuristics} we discuss sequential data gathering in the
context of adaptive design of AUV measurements.

We denote the data at $n_y$ locations by
$\by=(y_{T,1},y_{S,1},\ldots,y_{T,n_y},y_{S,n_y})^t$, representing the
temperature and salinity measurements gathered at the first round of
nodes in the grid. The conditional model for the data, given the true temperature
and salinity, is described by 
\begin{equation}\label{likelihood}
\by | \bxi \sim N( \bG \bxi, \bR), %\pi(\by | \bxi)= \frac{1}{(2\pi)^{m}|\bR|^{\frac{1}{2}}}e^{-\frac{1}{2}(\by -\bG\bxi)^T\bR^{-1}(\by-\bG\bxi)},
\end{equation}
where $\bG$ is a $2n_y \times 2n$ matrix containing '$1$'s on the
measurement indices of the grid and '$0$' otherwise. The size
$2n_y \times 2n_y$ covariance matrix $\bR$ contains diagonal entries
$r^2_T$ and $r^2_S$ which are measurement error variances of
temperature and salinity observations.

The GP mean and covariance in Eq. (\ref{prior}) are updated in the model when more information is available. The
updated distribution for temperature and salinity variables on the
spatial grid, given survey data $\by$, is Gaussian with mean and
covariance: 
\begin{eqnarray}\label{gp_upd}
  \bm &=& \bm(\by) = \bmu+\bSigma \bG^t (\bG \Sigma \bG^t+\bR)^{-1}(\by-\bG \bmu),  \\
  \bA &=& \bSigma - \bSigma \bG^t (\bG \bSigma \bG^t+\bR)^{-1} \bG
          \bSigma.\nonumber
\end{eqnarray}

@@

% The conditional mean $\bm$ will be important in the following
% derivation in Section \ref{sec:sur}.
The calculation of the design criterion in Eq. (\ref{sur}) requires
the marginal distribution of the data which is defined over $\bxi$ to
be $\by \sim N( \bG \bmu,\bG \bSigma \bG^t+\bR)$.  However, it turns
out that a simplified form of the calculation is possible because the
conditional mean in Eq (\ref{gp_hat}) is a linear (affine) function of
the data $\by$ and the covariance is not a function of the data.
Before inputting the data, the distribution of the conditional mean is
\begin{equation}\label{distxi} \bm \sim N(\bmu , \bPsi), \hspace{3mm}
  \bPsi=\bSigma \bG^t (\bG \bSigma \bG^t+\bR)^{-1} \bG \bSigma.
\end{equation} 
When considering only grid location $\bx$, we can
extract elements from its mean vector and covariance matrix to get the
bivariate distribution \begin{equation}\label{dist_mxi} 
  \bm(\bx) \sim N \left( \begin{bmatrix}
      \mu_{T}(\bx) \\
      \mu_{S}(\bx) \end{bmatrix}, \begin{bmatrix}
      \psi^2_{T}(\bx) & \psi_{T,S}(\bx)\\
      \psi_{T,S}(\bx) & \psi^2_{S}(\bx) \end{bmatrix} \right),
\end{equation}
which is important for the derivation of the closed-form solution of the design criteria.
